---
title: "Data Train | OT-SC-WS-04: Evaluating machine learning and artificial intelligence
  algorithms"
author:
- Max Westphal (max.westphal@mevis.fraunhofer.de)
output:
  html_document:
    df_print: paged
toc: yes
toc_depth: 3
toc_float: yes
editor_options: 
  markdown: 
    wrap: 80
---

# R session 2: Experimental design in ML performance evaluation

```{r}
## clear enviroment:
rm(list = ls())

## calc_costly determines if computationally expensive steps should be run during notebook rendering:
calc_costly <- FALSE

## save_vars determines if certain intermediate variables should be saved during notebook rendering:
save_vars <- FALSE
```

```{r packages, message=FALSE}
library(dplyr)
library(mlr3verse)
```

--------------------------------------------------------------------------------

## Exercise 1: Task preparation (15 minutes)

Today, we work with the **Cardiotocography Data Set** from the **UCI Machine
Learning Repository**. Let's import a preprocessed version of the data set.

```{r}
ctg_data <- readRDS("data/session2/ctg_data.rds")
dim(ctg_data)
head(ctg_data)
```

A basic documentation and the original data is provided at
<https://archive.ics.uci.edu/ml/datasets/cardiotocography>. There, it is stated
that

> 2126 fetal cardiotocograms (CTGs) were automatically processed and the
> respective diagnostic features measured. The CTGs were also classified by
> three expert obstetricians and a consensus classification label assigned to
> each of them. Classification was both with respect to a morphologic pattern
> (A, B, C. ...) and to a fetal state (N, S, P). Therefore the dataset can be
> used either for 10-class or 3-class experiments

We actually have a simplified version of the task with a binary target `status`
with two states `"suspect"`(positive class, including the original
`"pathological"` state) and `"normal"`. In addition, a few unneeded columns have
been removed from the original data set.

```{r}
summary(ctg_data$status)
```

Note: The data set has a `Date` variable, i.e. each observation has a time
stamp, which is rare for public data sets. The observations are actually sorted
according to `Date`. We will make use of this feature throughout this exercise.

```{r}
summary(ctg_data$Date)
```

```{r}
table(substr(ctg_data$Date, 1, 4))
```

The number of observations makes clear that we (strangely) have 56 observations
from 1980, a very small number compared to the remaining observations from the
years 1995-1998. One may reasonable suspect that these observations will be
quite distinct from the rest (or a coding error). Anyhow, we will remove the
observations from the data set for our purposes.

**Task: reduce `ctg_data` to only include observations from the year
1995-1998.**

```{r}
ctg_data <- ctg_data %>% filter(substr(ctg_data$Date, 1, 4) != "1980")
table(substr(ctg_data$Date, 1, 4))
```

Let's define the `mlr3` ML task for today:

```{r}
ctg_task <- as_task_classif(ctg_data %>% select(-Date),
                            target = "status",
                            positive = "suspect",
                            id = "ctg_task")
ctg_task
```

**Question: Why do we want exclude the `Date` variable from the feature set for
modeling purposes?**

-   A predictive model would be required to extrapolate (widely) for new, unseen
    data (with newer `Date` than those in `ctg_data)` which may lead to biased
    results

By the way, we will not cut of a distinct test set today (compared to session
1). The main reason is simplicity. If this was a real-life study, an additional
evaluation on a distinct test data set might be needed, depending on the
context.

```{r}
autoplot(ctg_task, type="duo")
```

--------------------------------------------------------------------------------

## Exercise 2: Common CV variants (15 minutes)

The focus in the following will be in defining different experimental designs
and observing and interpreting the different results. We would not do this in an
actual ML project but rather decide in advance which experimental design suits
the context of our study best, e.g. is best aligned with our study goals and/or
deployment scenarios.

### Cross-validation (10-fold)

We start by defining a very standard design, namely 10-fold CV.

```{r}
set.seed(1337)
exp_design_a <- rsmp("cv")
exp_design_a$instantiate(ctg_task)
exp_design_a$instance %>% plot()

```

### Cross-validation (5-fold)

**Task: define `exp_design_b` as 5-fold CV.**

```{r}
set.seed(1337)
exp_design_b <- rsmp("cv", folds=5)
exp_design_b$instantiate(ctg_task)
exp_design_b$instance %>% plot()
```

### Repeated cross-validation (5-fold, 4 repeats)

**Task: define `exp_design_c` as repeated CV with 5 folds and 4 repetitions.**

```{r}
set.seed(1337)
exp_design_c <- rsmp("repeated_cv", folds=5, repeats=4)
exp_design_c$instantiate(ctg_task)
exp_design_c$instance %>% plot()
```

--------------------------------------------------------------------------------

## Exercise 3: Custom experimental designs (15 minutes)

We have removed the `Date` variable for modeling purposes but we want to employ
it to define custom experimental designs in the following.

**Task: create a copy `ctg_data_ext` of `ctg_data` with one additional variable
showing the year component of `Date`**.

```{r}
ctg_data_ext <- ctg_data %>% mutate(Year = substr(Date, 1, 4))
table(ctg_data_ext$Year)
```

### Cross-validation (grouped by year)

Usual (random) cross-validation assumes a homogeneous data sample and assesses
generalization to new (unseen) samples. We might suspect that patient
characteristics change over time. Thus, for the next, experimental design we
want to access how models generalize to different years (unseen during
training). Have a close look at the (special) syntax, you might need it quite
soon.

```{r}
exp_design_d <- rsmp("custom_cv")
exp_design_d$instantiate(ctg_task, f=as.factor(ctg_data_ext$Year))
exp_design_d$instance %>% str()

```

### Custom design "train 1, predict 1"

Now we want to define a `"custom"` design, where data from one year is used for
training and assessed on unseen data from the following year.

```{r}
rsmp("custom")
```

```{r}
unique(ctg_data_ext$Year)
```

**Task: define two lists`train_sets_e` and `test_sets_e` . Both lists should
have length 3. (Why?) Each element is a vector of row indies defining train and
test sets as indicated above.**

(Hint: you need to incorporate the variable `ctg_data_ext$Year` )

```{r}
train_sets_e <- list(
  which(ctg_data_ext$Year == "1995"),
  which(ctg_data_ext$Year == "1996"),
  which(ctg_data_ext$Year == "1997")
)
test_sets_e <- list(
  which(ctg_data_ext$Year == "1996"),
  which(ctg_data_ext$Year == "1997"),
  which(ctg_data_ext$Year == "1998")
) 
```

```{r}
exp_design_e <- rsmp("custom")
exp_design_e$instantiate(ctg_task,
                         train_sets = train_sets_e,
                         test_sets = test_sets_e)
exp_design_e$instance %>% str()
```

### Custom design "train 2, predict 1"

Lets slightly adapt the last design and now train on the two years preceding the
year before the 'test year'.

**Task: define two lists`train_sets_f` and `test_sets_f` . Both lists should
have length 2. (Why?) Each element is a vector of row indies defining train and
test sets as indicated above.**

```{r}
train_sets_f <- list(
  which(ctg_data_ext$Year %in% c("1995", "1996")),
  which(ctg_data_ext$Year %in% c("1996", "1997"))
)
test_sets_f <- list(
  which(ctg_data_ext$Year == "1997"),
  which(ctg_data_ext$Year == "1998")
) 
```

```{r}
exp_design_f <- rsmp("custom")
exp_design_f$instantiate(ctg_task,
                         train_sets = train_sets_f,
                         test_sets = test_sets_f)
exp_design_f$instance %>% str()
```

--------------------------------------------------------------------------------

## Exercise 4: Model tuning (10 minutes)

Now that we defined all these different experimental design (training/validation
splits), lets compare the benchmark results.

Today, we'll be employing a famous and very powerful algorithm extreme gradient
boosting (XGBoost). This will get computationally intensive, at least much more
than before. We can use the parallelization capabilities of `mlr3` to reduce the
run time of our benchmark experiment.

**Note: The code blocks below (`xgb_learner_tuned_xyz <- tune(…)` ) may take a
long time to run. We have pre-computed the results for you. They will be loaded
at the start of the next section. Make sure that you understand what's happening
in the next code chunks even if you don't run them.**

```{r}
future::availableCores()
num_threads <- round(0.8*future::availableCores())
# num_threads <- 1 # when working on RStudio Server
```

```{r}
future::plan("multisession")
```

To get a reasonable default hyperparameter tuning space we make use of the
`mlr3tuningspaces` package.

```{r}
ts <- mlr3tuningspaces::lts("classif.xgboost.default")
xgb_learner <- ts$get_learner()
xgb_learner$predict_type <- "prob"
set_threads(xgb_learner, n = num_threads)
xgb_learner

```

Lets sample 40 hyperparameters from this space now.

```{r}
set.seed(1337)
num_hp <- 40
xgb_space <- xgb_learner$param_set$search_space()
xgb_hp <- generate_design_lhs(xgb_space, num_hp)
xgb_hp
```

**Question: What is the function `generate_design_lhs` doing?**

-   See `?generate_design_lhs` and also
    <https://en.wikipedia.org/wiki/Latin_hypercube_sampling>

```{r}
if(save_vars){
  saveRDS(xgb_hp, "data/session2/xgb_hp.rds")
}
xgb_hp <- readRDS("data/session2/xgb_hp.rds")
```

```{r, eval=calc_costly, results='hide'}
set.seed(1337)
xgb_learner_tuned_a <- mlr3tuning::tune(
  method = "design_points",
  task = ctg_task,
  learner = xgb_learner,
  resampling = exp_design_a,
  measures = msrs(c("classif.auc")),
  design = xgb_hp$data,
  batch_size = num_threads
)
```

--------------------------------------------------------------------------------

```{r, eval=calc_costly, results='hide'}
set.seed(1337)
xgb_learner_tuned_b <- mlr3tuning::tune(
  method = "design_points",
  task = ctg_task,
  learner = xgb_learner,
  resampling = exp_design_b,
  measures = msrs(c("classif.auc")),
  design = xgb_hp$data,
  batch_size = num_threads
)
```

```{r, eval=calc_costly, results='hide'}
set.seed(1337)
xgb_learner_tuned_c <- mlr3tuning::tune(
  method = "design_points",
  task = ctg_task,
  learner = xgb_learner,
  resampling = exp_design_c,
  measures = msrs(c("classif.auc")),
  design = xgb_hp$data,
  batch_size = num_threads
)
```

```{r, eval=calc_costly, results='hide'}
set.seed(1337)
xgb_learner_tuned_d <- mlr3tuning::tune(
  method = "design_points",
  task = ctg_task,
  learner = xgb_learner,
  resampling = exp_design_d,
  measures = msrs(c("classif.auc")),
  design = xgb_hp$data,
  batch_size = num_threads
)
```

```{r, eval=calc_costly, results='hide'}
set.seed(1337)
xgb_learner_tuned_e <- mlr3tuning::tune(
  method = "design_points",
  task = ctg_task,
  learner = xgb_learner,
  resampling = exp_design_e,
  measures = msrs(c("classif.auc")),
  design = xgb_hp$data,
  batch_size = num_threads
)
```

```{r, eval=calc_costly, results='hide'}
set.seed(1337)
xgb_learner_tuned_f <- mlr3tuning::tune(
  method = "design_points",
  task = ctg_task,
  learner = xgb_learner,
  resampling = exp_design_f,
  measures = msrs(c("classif.auc")),
  design = xgb_hp$data,
  batch_size = num_threads
)
```

```{r, eval=calc_costly}
xgb_learner_tuned_all <- list(xgb_learner_tuned_a,
                              xgb_learner_tuned_b,
                              xgb_learner_tuned_c,
                              xgb_learner_tuned_d,
                              xgb_learner_tuned_e,
                              xgb_learner_tuned_f)
xgb_lta_archives <- 
  lapply(xgb_learner_tuned_all, function(x) as.data.table(x$archive))
names(xgb_lta_archives) <- letters[1:6]

```

--------------------------------------------------------------------------------

## Exercise 5: Comparison of evaluation results based on different designs (20 minutes)

In this section, we start the (most important) interpretation part. Recall that,
so far, we have used six different experimental setup on the same ML task. Now,
we will have a look at how these different designs affect our evaluation
results.

Let's have a look at the archives attribute from all benchmark results. Load the
pre-computed results unless you were brave enough to run the training/tuning
chunks yourself.

```{r}
if(save_vars){
  saveRDS(xgb_lta_archives, "data/session2/xgb_lta_archives.rds")
}
xgb_lta_archives <- readRDS("data/session2/xgb_lta_archives.rds")
```

```{r}
str(xgb_lta_archives, 1)
```

Of course, we are most interested in the AUCs.

```{r}
xgb_lta_aucs <- sapply(xgb_lta_archives, function(x) x$classif.auc) %>% as.data.table()
head(xgb_lta_aucs)
```

```{r}
pairs(xgb_lta_aucs)
```

**Question: How can the pairs plot be interpreted?**

-   One dot represents one hyperparameter combination and its performance under
    different experimental design. While everything is positively correlated, we
    can the the systematic differences due to the different experimental
    designs. In particular design A-C are much more strongly correlated compared
    to the other designs.

**Task: Estimate and interpret the correlation matrix of the `aucs` object.**

```{r}
cov(xgb_lta_aucs) %>% cov2cor()
```

**Task: For each experimental design, find the hyperparameter configuration
which maximizes the AUC and save the index in the vector `best`.**

```{r}
apply(xgb_lta_aucs, 2, max)
best <- apply(xgb_lta_aucs, 2, which.max)
best
```

Let's have a look at the corresponding hyperparameters.

```{r}
xgb_hp$data[unique(best), ]
```

**Task: Fill in the following table, describing the context and resources (no R
code required, only common sense). Which design(s) are most appropriate?**

+---+--------------+-----------+-----------------------------------+-----------+
|   | Experimental | R         | Context\                          | R         |
|   | design       | esources\ | (what are we estimating?)         | easonable |
|   |              | (training |                                   | design?   |
|   |              | runs per  |                                   |           |
|   |              | HP)       |                                   |           |
+===+==============+===========+===================================+===========+
| A | 10-fold CV\  | 10        | expected AUC when training models | probably  |
|   |              |           | based on 90% of sample size       | not       |
+---+--------------+-----------+-----------------------------------+-----------+
| B | 5- fold CV\  | 5         | expected AUC when training models | probably  |
|   |              |           | based on 80% of sample size       | not       |
+---+--------------+-----------+-----------------------------------+-----------+
| C | 5-fold CV\   | 20        | expected AUC when training models | probably  |
|   | (4           |           | based on 80% of sample size       | not       |
|   | repetitions) |           |                                   |           |
+---+--------------+-----------+-----------------------------------+-----------+
| D | custom block | 4         | expected AUC when training models | probably  |
|   | design       |           | based on various fractions of     | not       |
|   | (years)      |           | sample size from different years  |           |
+---+--------------+-----------+-----------------------------------+-----------+
| E | custom       | 3         | expected AUC when training models | maybe     |
|   | design       |           | based on samples from past year   |           |
|   | "train 1,    |           |                                   |           |
|   | predict 1"   |           |                                   |           |
+---+--------------+-----------+-----------------------------------+-----------+
| F | custom       | 2         | expected AUC when training models | maybe     |
|   | design       |           | based on samples from past two    |           |
|   | "train 2,    |           | years                             |           |
|   | predict 1"   |           |                                   |           |
+---+--------------+-----------+-----------------------------------+-----------+

**Question: For design D, E and F, do the different sample sizes over the years
pose a problem?**

-   Sort of. The performances differences can not be attributed solely to the
    different time periods used for training and prediction.

**Question: Can you think of other relevant custom design for this specific
task?**

-   A remedy for the issue mentioned above might be to not split based on
    calendar years but rather based on time such that all time windows have an
    equal number of observations.

--------------------------------------------------------------------------------

## Exercise 6: Nested CV including inference (30 minutes)

The goal for this exercise will be to conduct statistical inference for a more
complex experimental design, namely nested cross-validation. First, let's have a
look at the documentation of the `tune_nested` function.

```{r, eval=FALSE}
?tune_nested
```

```{r, eval=calc_costly, results='hide'}
set.seed(1337)
xgb_learner_tuned_nested <- mlr3tuning::tune_nested(
  method = "random_search",
  task = ctg_task,
  learner = xgb_learner,
  inner_resampling = rsmp("holdout", ratio=0.8),
  outer_resampling = rsmp("cv", folds = 5),
  measure = msr("classif.auc"),
  term_evals = 40,
  batch_size = num_threads
)
xgb_ltn_preds <- as.data.table(xgb_learner_tuned_nested$prediction())
xgb_ltn_resamp <- xgb_learner_tuned_nested$resampling$instance
```

```{r}
if(save_vars){
  saveRDS(xgb_ltn_preds, "data/session2/xgb_ltn_preds.rds")
  saveRDS(xgb_ltn_resamp, "data/session2/xgb_ltn_resamp.rds")
}
xgb_ltn_preds <- readRDS("data/session2/xgb_ltn_preds.rds")
xgb_ltn_resamp <- readRDS("data/session2/xgb_ltn_resamp.rds")
```

Let's have a look at parts of the object.

```{r}
xgb_ltn_preds 
```

```{r}
xgb_ltn_resamp
```

Finally, let's utilize the `ci.cvAUC` function from the `cvAUC` package to
estimate the final AUC (outer CV loop) and also calculate a confidence interval.

```{r, message=FALSE}
library(cvAUC)
```

```{r, eval=FALSE}
?ci.cvAUC
```

**Task: apply the `ci.cvAUC` function to calculate confidence intervals for the
final AUC.**

```{r}
ci.cvAUC(
  predictions = xgb_ltn_preds$prob.suspect, 
  labels = xgb_ltn_preds$truth,
  folds = xgb_ltn_resamp %>% arrange(row_id) %>% {as.integer(.$fold)}
)
```

**Question: what is the advantage of nested CV (as conducted above) compared to
usual 5-fold CV (as defined earlier in experimental design B)?**

-   Reduced bias for the outer loop AUC estimate because model selection (inner
    loop) and performance estimation are now clearly separated in contrast to
    conventional CV (both based on outer/only loop).

**Question: For what quantity have we just calculated an estimate and confidence
interval?**

-   The expected AUC of models trained via the XGBoost algorithm (applied to 80%
    of the data set size, i.e. 1656 observations) whereby the hyperparameter
    have been optimized in the inner CV loop via random search. (Keep in mind
    that our target population is from years 1995 to 1998 and )

**Question: Connecting these results to exercise 5, how relevant is our nested
CV result from a practical viewpoint? Can the relevance be improved?**

-   Probably more relevant custom nested CV variant would be to define outer and
    inner loop based on the Date variable, i.e. in both loops use earlier
    samples for training and later samples for prediction.

**Question: If we wanted to implement a trained predictive model in (clinical)
practice - what would we need to do?**

-   Apply the XGBoost algorithm with the 'winning' hyperparameter combination
    (as determined by the inner nested CV loop) to the complete data set.

--------------------------------------------------------------------------------

## Exercise 7: Tackle your own ML evaluation task (optional)

Congratulations - you have reached the end of the course exercises.

Feel free to use any remaining time for your own ML evaluation problem, i.e. try
to narrow down important properties of your evaluation study such as

-   ML task properties, in particular the baseline/benchmark and the evaluation
    metric(s)

-   Benchmark experimental design(s)

-   Statistical analysis for performance evaluation
